On the Minimum Core Mass for Giant Planet Formation

### 2013-06-20 tex/draft/draft2.tex
updated analytic section

We consider a two-layer atmosphere structure model as presented in section \ref{model}. We approximate the outer radiative zone of the atmosphere as nearly isothermal. Moreover, we do not take into account the self-gravity of the atmosphere in either region. We derive the structure of an isothermal non self-gravitating atmosphere is section \ref{iso}.

In the next subsections we derive the structure and cooling behavior of a two-layer non self-gravitating atmosphere, deferring some of the calculations to \App{sec:analytic}. We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}. Finally, we derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.

We now consider the structure of a low mass (non self-gravitating) isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}). From equation (\ref{eq:structa}) and without taking into account the self-gravity of the atmosphere, the resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
\end{equation} 
where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius. In our study, $n_{\rm{fit}}=R_{\rm H}/R_{\rm B}$.

\noindent $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.  If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\cb}$.   With this assumption, we get a constant value for $\nabla_{\rm d}$.  As $\delrad=d \ln T/d \ln P$, the temperature profile integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1+\alpha} - 1 \right] \, ,
\end{equation} 
where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
We  apply \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there $P\cb$ satisfies the relation $P\cb \gg P_{\rm d}$. This is motivated by the fact that the density, and hence pressure profile, falls exponentially (see equation \ref{eq:rhoiso}). The resulting temperature contrast at the radiative-convective boundary is
\begin{equation}\label{eq:Tcb}
\chi \equiv {T\cb\over T_{\rm d}} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
\end{equation} 
For  $\alpha=0$ and $\beta =2$,  and assuming $\delad = 2/7$, we find $\chi \approx 1.5$. Values of $\chi$ and $\nabla_\infty$ for other choices of $\beta$ are summarized in \App{sec:analytic}.

 We now eliminate $\nabla\di$ from equation (\ref{eq:Pcbapprox}) to obtain a relation between temperature and pressure as a function of the pressure at the radiative-convective boundary $P\cb$. From equation (\ref{eq:radTP}), it follows that
 \begin{equation}\label{eq:TP}
{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1+\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1+\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
\end{equation} 
 We can then determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation (\ref{eq:struct a}) as 
\begin{equation}\label{eq:RCBint}
{R_B \over R\cb} = \int_{P\di}^{P\cb} {T \over T_{\rm d}} {dP \over P}\, .
\end{equation} 
An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However using \Eq{eq:TP} in the integral gives
\begin{equation}\label{eq:Rcb}
{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
\end{equation} 
with an extra correction term, $\theta < 1$.  The form we chose for the correction term allows us to relate the disk and radiative-convective boundary pressures as :
 \begin{equation}\label{eq:PcbRcb}
P\cb = \theta P_{\rm d} e^{R_B/R\cb} \, .
\end{equation}   
In the $P\cb \gg P_{\rm d}$ limit, the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$. Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal. For $\delad=2/7$, $\alpha=0$ and $\beta=2$, we find $\theta \approx 0.556$. Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.  A simple analytic expression for $\theta$ is not possible.

\subsection{Simplified cooling model}
\label{coolingan}
We now use a simplified version of the cooling model introduced in section \ref{cooling} in order to obtain an analytic cooling model. As before, we assume a polytropic equation of state and neglect self-gravity.

\subsubsection{Luminosity,  Energy \& Mass}
From equation (\ref{eq:delradan}), the luminosity that emerges at the radiative-convective boundary is
\begin{equation} \label{eq:Lcb}
L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}, % \left({P\di \over P\cb}\right)%^{1+\alpha}
\end{equation} 
where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and define 
\begin{equation} 
L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}, 
\end{equation} 
with $T = \chi T_{\rm d}$ at the radiative-convective boundary.

@@

To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at the core mass $M\co$.  The density profile of an adiabatic atmosphere follows from hydrostatic balance as 
\begin{equation}\label{eq:rhoconv} 
\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over R\cb}  \right]^{1/(\gamma -1)},
\end{equation} 
where we define an effective Bondi radius,
\begin{equation}
R_B' \equiv {G M\co \over C_{\rm P} T\cb} = {\delad \over \chi} R_B
\end{equation} 
to simplify expressions, with $C_{\rm P}$ the specific heat at constant pressure.

  Deep in the atmosphere, where $r \ll R\cb \lesssim R_B'$ the density profile is $\rho \propto r^{-1/(\gamma -1)}$.  Since the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$, only polytropes with $\gamma < 3/2$ (i.e. $\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.  We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.

The total (thermal and gravitational) energy in an adiabatic atmosphere could be evaluated from the virial theorem (see section \ref{cooling}).  More simply, we use the result for temperature profiles in deep (but non-self-gravitating) convective regions:
\begin{equation}
T \approx {G M\co \over C_P r} = T\cb {R_B' \over r}\, .
\end{equation} 
The internal energy per unit mass of an ideal gas is thus $u = C_V T = (1 - \delad) G M\co/r$ and the specific energy  deep in the atmosphere is
\begin{equation}
e = e_g + u = -\delad {GM\co \over r} \, .
\end{equation} 
Thus the total energy for $\gamma < 3/2$ is thus
\begin{eqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{r\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \\
&\approx& - 8 \pi P\cb {R_B'^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {(R_B / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {R_B'^{7/2} \over \sqrt{R\co}}
\end{eqnarray} 
where the final expression takes $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.

The mass of the adiabatic atmosphere is given by
\begin{eqnarray} 
\label{eq:Matman}
M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb}, % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
\end{eqnarray}
in the limit $R\co \ll R\cb \ll R_B'$. From equation (\ref{eq:MatmISO}) and replacing $R_{\rm c}$ with $R\cb$, the mass in the isothermal region is given by $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.

We can eliminate $R\cb$ using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
\begin{equation} \label{eq:crit}
{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
\end{equation} 
where we introduce a characteristic pressure
\begin{equation} 
P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {R_B'}^4}\, .
%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {R_B'}^4}\, .
\end{equation} 
For the atmosphere to initiate unstable gas accretion, $M_{\rm atm} = M\co$. We thus require
\begin{equation} \label{eq:Pcbc}
P\cb = \xi P_M
\end{equation} 
where  the logarithmic factor
\begin{equation}\label{eq:xi}
\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
\end{equation} 
is found by numerically solving the above transcendental equation. The physical solution has $\xi >1$, but is typically of order unity. Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.
%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.

\subsubsection{Cooling Time}
Neglecting the surface terms from equation (\ref{eq:coolingglobal}), the total time to cool the atmosphere from an initially fully adiabatic state is \footnote{The relevance of the neglected surface terms is discussed in \App{sec:analytic}}
\begin{eqnarray} 
t_{\rm  cool} &=& -\int {d E \over L} = - \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, ,
%&\approx& {0.16 }{P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}}
\end{eqnarray} 
where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.

We define the timescale for runaway gas accretion to initiate, when $M_{\rm{atm}}=M_{\rm c}$, as the \textit{crossover time} $t_{\rm{co}}$. This time is found from \Eqs{eq:tcool}{eq:Pcbc} to be
\begin{eqnarray} 
\label{eq:tcoolan}
t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(a\over 10\right)^{15 \over 14}} \yr
\end{eqnarray} 
for $\beta = 2$ and $\mc \equiv M_{\rm c} / M_{\oplus}$.  
This timescale is too long compared to the average lifetime of a protoplanetary disk --- this is likely due to our neglect of self-gravity.  Nevertheless, the trends are informative.   A reduction in opacity gives faster cooling, as long as the optically thick assumption holds.  Lower disk temperatures also decrease the crossover time;  even though higher temperatures result in higher luminosities, they also give higher dust opacities.   For different $\beta$ values, the temperature dependence is $t_{\rm co} \propto F_T^{ \beta + 1/2}$.  The cooling timescale only weakly depends on disk mass or pressure, via the  logarithmic factor $\xi$.  This weak dependence is a consequence of cooling coming from the radiative-convective boundary.  Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity. % and is here given the appropriate value for the nominal parameters.

\subsubsection{Critical Core Mass}\label{sec:critmass}
%We use the results of the previous section to derive the minimum core mass that can form a giant planet by gas accretion.  Since our structure calculations do not include the self-gravity of the atmosphere, we simply assume that runaway growth occurs once the atmosphere's mass (as calculated in the absence of self-gravity) reaches the core mass, $M_{\rm atm} \ge M_{\rm c}$.  This approximation was used by R06 and is consistent with detailed evolution models (refs).

We can define a \textit{critical more mass} as the core mass for which $M_{\rm{atm}}=M_{\rm c}$ within a typical gas disk lifetime:
\begin{equation}
t_{\rm co} = 3 \times 10^6 ~{\rm yr} \,.
\end{equation} 
From equation (\ref{eq:tcoolan}), the resulting critical core mass is
\begin{equation}\label{eq:Mcrit}
M_{\rm crit} \approx 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \left(\au \over 10\right)^{9 \over 14}} \; M_\oplus.
\end{equation} 
Similarly to the crossover time, the critical mass is too large due to the neglect of self-gravity.  The scaling with disk properties is similar: all quantities are raised to the $3/5$ power.

@@

>>>>

We consider a two-layer atmosphere structure model as presented in section \ref{model}, and we make two key assumptions: we neglect the self-gravity of the atmosphere, and we assume that the outer radiative region is very thick and hence the pressure at the radiative-convective boundary is much larger than the disk pressure $P\di$. We will show in section \ref{KH} that these assumptions limit the applicability of this simplified model during the late and early stages of the atmosphere evolution, respectively. We also approximate the structure of the radiative zone as nearly isothermal, and note that this assumption is consistent with the numerical solutions (see section \ref{profiles}).

As self-gravity is quantitatively important, we expect our analytic model to only provide scalings for the full numerical model rather than absolute values.

%We approximate the outer radiative zone of the atmosphere as nearly isothermal. Moreover, we do not take into account the self-gravity of the atmosphere in either region. We derive the structure of an isothermal non self-gravitating atmosphere is section \ref{iso}.

We use the quasi-static model of section \ref{sec2} to derive expressions for the mass, energy and luminosity of a static non self-gravitating atmosphere. We use this to derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}. We defer the details of our calculations to \App{sec:analytic}.

%In the next subsections we derive the structure and cooling behavior of a two-layer non self-gravitating atmosphere, deferring some of the calculations to \App{sec:analytic}. We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}. Finally, we derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.

%\subsection{Simplified cooling model}
%\label{coolingan}

%In this section we

%We now use a simplified version of the cooling model introduced in section \ref{cooling} in order to obtain an analytic cooling model. As before, we assume a polytropic equation of state and neglect self-gravity.

\subsection{Mass, Energy and Luminosity}
\label{MELan}

In this section, we derive simplified analytic expressions for the mass, energy and luminosity of the atmosphere.

Our assumption of a nearly isothermal radiative region yields simple expressions for the temperature and pressure at the radiative-convective boundary as a function of the nebular temperature and pressure. These expressions will helps us estimate the mass and energy of the atmosphere. We account for the departure from a perfectly isothermal layer (i.e., constant temperature and exponential pressure profile, see also equation \ref{eq:rhoiso}) by introducing two correction terms $\chi$ and $\theta$ such that

\begin{equation}
\label{eq:Tcb}
T\cb = \chi T\di
\end{equation} 
and
 \begin{equation}\label{eq:PcbRcb}
P\cb = \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, ,
\end{equation}
where $\chi$ can be approximated as
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} ,
\end{equation}
with $\nabla_{\infty}=(1+\alpha)/(4-\beta)$ the radiative temperature gradient as $T, P \rightarrow \infty$. A simple analytic expression for $\theta$ is not possible. Both $\chi$ and $\theta$ depend on $\alpha$, $\beta$ and the adiabatic gradient $\delad$. For $\alpha=0$, $\beta=2$ and $\delad=2/7$, we find $\chi \approx 1.5$ and $\theta \approx 0.556$. Values for $\theta$ and $\chi$ for other choices of $\beta$ are summarized in \App{sec:analytic}, Table I.

We then determine the density profile in the convective region of the atmosphere to estimate the mass of the atmosphere. Integrating the equation of hydrostatic balance (\ref{eq:structa}) and holding the mass fixed at $M\co$, we find

\begin{equation}\label{eq:rhoconv} 
\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over R\cb}  \right]^{1/(\gamma -1)},
\end{equation} 
where we define an effective Bondi radius,
\begin{equation}
R_B' \equiv {G M\co \over C_{\rm P} T\cb} = {\delad \over \chi} R_B
\end{equation} 
to simplify expressions, with $C_{\rm P}$ the specific heat at constant pressure.

From this we can calculate the mass of the convective region as

\begin{eqnarray} 
\label{eq:Matman}
M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
&\approx& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb}, % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
\end{eqnarray}
in the limit $R\co \ll R\cb \ll R_B'$ and where $\gamma=7/5$ in the final expression. The mass in the isothermal region can be estimated as $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_{\rm B}$ (see \App{sec:analytic} for a derivation).  We see that $M_{\rm{iso}} \ll M_{\rm{atm}}$ and therefore it can be neglected.

The total (thermal and gravitational) energy in an adiabatic atmosphere can be evaluated from the virial theorem (see section \ref{cooling}).  More simply, we use the result for temperature profiles in deep, but non-self-gravitating, convective regions:
\begin{equation}
T \approx {G M\co \over C_P r} = T\cb {R_B' \over r}\, .
\end{equation} 
The internal energy per unit mass of an ideal gas is thus $u = C_V T = (1 - \delad) G M\co/r$ and the specific energy  deep in the atmosphere is
\begin{equation}
e = e_g + u = -\delad {GM\co \over r} \, .
\end{equation} 
The total energy follows as
\begin{eqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{r\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
&\approx& - 8 \pi P\cb {R_B'^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {(R_B / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {R_B'^{7/2} \over \sqrt{R\co}}
\end{eqnarray} 
when $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.

We note that the approximation in the energy expression of equation (\ref{eq:Ean}) is valid only deep in the adiabatic atmosphere, where  $\rho \propto r^{-1/(\gamma -1)}$ from equation (\ref{eq:rhoconv}).  As the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$, only polytropes with $\gamma < 3/2$ (i.e. $\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.  Equation (\ref{eq:Ean}) is therefore only valid when $\gamma<3/2$. % We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.

Finally, we can estimate the luminosity that emerges at the radiative-convective boundary from equation (\ref{eq:structc}) assuming $\delrad=\delad$ at the radiative-convective boundary, as

\begin{equation} \label{eq:Lcb}
L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}, % \left({P\di \over P\cb}\right)%^{1+\alpha}
\end{equation} 
where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and define 
\begin{equation} 
L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}, 
\end{equation}

@@

%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.

\subsection{Simplified Cooling Model}
\label{coolingan}

We use the equations obtained in section \ref{MELan} to estimate the atmosphere mass $M_{\rm{atm}}$ for which $M_{\rm{atm}} \sim M\co$ and runaway accretion commences. We then use a simplified version of the cooling model developed in section \ref{cooling} to estimate the time elapsed until collapse.

We can eliminate $R\cb$ from equation (\ref{eq:Matman}) using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
\begin{equation} \label{eq:crit}
{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
\end{equation} 
where we introduce a characteristic pressure
\begin{equation} 
P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {R_B'}^4}\, .
%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {R_B'}^4}\, .
\end{equation} 
For $M_{\rm atm} = M\co$ in equation (\ref{eq:crit}) we require
\begin{equation} \label{eq:Pcbc}
P\cb = \xi P_M ,
\end{equation} 
where  the logarithmic factor
\begin{equation}\label{eq:xi}
\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
\end{equation} 
is found by numerically solving the above transcendental equation. The physical solution has $\xi >1$, but is typically of order unity. Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.

If we neglect the surface terms from equation (\ref{eq:coolingglobal}) \footnote{The relevance of the neglected surface terms is discussed in \App{sec:analytic}}, the total time elapsed until $M_{\rm{atm}}= M\co$ is given by

\begin{eqnarray} 
t_{\rm  cool} &=& -\int {d E \over L} = - \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, .
%&\approx& {0.16 }{P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}}
\end{eqnarray} 
%where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.

We define the timescale for runaway gas accretion to initiate, when $M_{\rm{atm}}=M_{\rm c}$, as the \textit{crossover time} $t_{\rm{co}}$. This time is found from \Eqs{eq:tcool}{eq:Pcbc} to be
\begin{eqnarray} 
\label{eq:tcoolan}
t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(a\over 10\right)^{15 \over 14}} \yr
\end{eqnarray} 
for $\beta = 2$ and $\mc \equiv M_{\rm c} / M_{\oplus}$.  
This timescale is too long compared to the average lifetime of a protoplanetary disk. We will show in section \ref{KH} that this is due to our neglect of self-gravity in the analytic model.  Nevertheless, the trends in the atmosphere evolution predicted by our analytic model are informative.  A reduction in opacity accelerates cooling as long as the optically thick assumption holds.  Lower disk temperatures also decrease the crossover time --- even though higher temperatures result in higher luminosities, they also result in higher dust opacities.   For different $\beta$ values, the crossover time depends on the disk temperature as $t_{\rm co} \propto F_T^{ \beta + 1/2}$.  The cooling timescale depends on the disk pressure only weakly via the  logarithmic factor $\xi$.  We show in section \ref{TPeffects} that the numerical model results in a qualitatively similar dependence on disk temperature and pressure.  %This weak dependence is a consequence of cooling coming from the radiative-convective boundary.  Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity. % and is here given the appropriate value for the nominal parameters.

\subsection{Critical Core Mass}\label{sec:critmass}
%We use the results of the previous section to derive the minimum core mass that can form a giant planet by gas accretion.  Since our structure calculations do not include the self-gravity of the atmosphere, we simply assume that runaway growth occurs once the atmosphere's mass (as calculated in the absence of self-gravity) reaches the core mass, $M_{\rm atm} \ge M_{\rm c}$.  This approximation was used by R06 and is consistent with detailed evolution models (refs).

We use the results of section \ref{coolingan} to estimate the core mass for which  runaway accretion commences, i.e. $M_{\atm}=M\co$, within a typical lifetime of a protoplanetary disk: 
\begin{equation}
t_{\rm co} = 3 \times 10^6 ~{\rm yr} \,.
\end{equation} 
We denote this core mass as the \textit{critical core mass} $M_{\rm crit}$. From equation (\ref{eq:tcoolan}), $M_{\rm{crit}}$ is given by
\begin{equation}\label{eq:Mcrit}
M_{\rm crit} \approx 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \left(\au \over 10\right)^{9 \over 14}} \; M_\oplus.
\end{equation}

Similarly to the crossover time, the critical mass is too large due to our neglect of self-gravity.  The scaling with disk properties is similar: all quantities are raised to the $3/5$ power.

@@

We consider the structure of a low mass non self-gravitating isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}). From equation (\ref{eq:structa}) the resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho_{\rm d} \exp \left({R_{\rm B} \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_{\rm B} \over r  \right),
\end{equation} 
where the approximate inequality is valid deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius.

\noindent The mass $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.  If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\cb}$.   With this assumption, we get a constant value for $\nabla_{\rm d}$.  As $\delrad=d \ln T/d \ln P$, the temperature profile integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1+\alpha} - 1 \right] \, ,
\end{equation} 
where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
By applying \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there $P\cb$ satisfies the relation $P\cb \gg P_{\rm d}$, we end up with the expression in equation (\ref{eq:Tcb}), $T\cb=\chi T\di$, with $\chi$ defined in equation (\ref{eq:chi}).

 . %\begin{equation}\label{eq:Tcb}
%\chi \equiv {T\cb\over T_{\rm d}} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
%\end{equation} 
%For  $\alpha=0$ and $\beta =2$,  and assuming $\delad = 2/7$, we find $\chi \approx 1.5$. Values of $\chi$ and $\nabla_\infty$ for other choices of $\beta$ are summarized in \App{sec:analytic}.

 We now eliminate $\nabla\di$ from equation (\ref{eq:Pcbapprox}) to obtain a relation between temperature and pressure as a function of the pressure at the radiative-convective boundary $P\cb$. From equation (\ref{eq:radTP}), it follows that
 \begin{equation}\label{eq:TP}
{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1+\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1+\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
\end{equation} 
 We can then determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation (\ref{eq:structa}) as 
\begin{equation}\label{eq:RCBint}
{R_B \over R\cb} = \int_{P\di}^{P\cb} {T \over T_{\rm d}} {dP \over P}\, .
\end{equation} 
An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However, using \Eq{eq:TP} in the integral gives
\begin{equation}\label{eq:Rcb}
{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
\end{equation} 
with an extra correction term, $\theta < 1$.  From this we arrive at the relation between $P\cb$ and $P\di$ given by equation (\ref{eq:PcbRcb}).

% The form we chose for the correction term allows us to relate the disk and radiative-convective boundary pressures as :
% \begin{equation}\label{eq:PcbRcb}
%P\cb = \theta P_{\rm d} e^{R_B/R\cb} \, .
%\end{equation}   
%In the $P\cb \gg P_{\rm d}$ limit, the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$. Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal. For $\delad=2/7$, $\alpha=0$ and $\beta=2$, we find $\theta \approx 0.556$. Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.  A simple analytic expression for $\theta$ is not possible.
